Automated reading of residual plots with computer vision models

### 2024-01-22 paper/paper.tex
update paper methodology and toc

@@

There is a substantial need to develop an approach that alleviates
people's workload by automating repetitive tasks and providing
standardized results in a controlled environment. The large-scale
evaluation of lineups is impractical without the use of technology and
machines.

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatterplot diagnostics. \citet{wilkinson2005graph} expanded
on this work, introducing graph theoretic scagnostics, which encompassed
nine computable measures applied to planar proximity graphs. These
measures, including ``Outlying'', ``Skinny'', ``Stringy'', ``Straight'',
``Monotonic'', ``Skewed'', ``Clumpy'', and ``Striated'' aimed to
characterize outliers, shape, density, trend, and coherence of the data.
While this approach has been inspiring, there is a recognition
\citep{buja2009statistical} that it may not capture all the necessary
visual features that differentiate actual residual plots from null
plots. A more promising alternative entails enabling computers to learn
the function for extracting visual features from residual plots.
Essentially, this means empowering computers to discern the crucial
visual features for residual diagnostics and determining the method to
extract them. Modern computer vision models are well-suited for
addressing this challenge.

@@

@@

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for evaluating residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure (SSIM) (ref here). The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}.

@@

The distance metrics proposed in this paper takes into account the fact
that we try to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted model is
from a correctly specified model. For classical normal linear regression
model, residuals are derived from the fitted values
\(\hat{\boldsymbol{y}}\) and observed values \(\boldsymbol{y}\).
However, if the data generating process is known and the model is
correctly specified, by the Frisch-Waugh-Lowell theorem (ref here),
residuals \(\boldsymbol{e}\) can be written as a linear transformation
of the error \(\boldsymbol{\varepsilon}\):

\begin{equation}
\label{eq:residual-1}
\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon},
\end{equation}

\noindent where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator, \(\boldsymbol{X}\) is the design matrix,
\(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity matrix, and \(n\) is
the number of observations.

One of the assumptions of the classical normal linear regression model
is the error \(\boldsymbol{\varepsilon}\) follows a multivariate normal
distribution with zero mean and constant variance, i.e.,
\(N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\). Using equation
\ref{eq:residual-1}, it can be known that residuals \(\boldsymbol{e}\)
also follow a certain probability distribution, which will be denoted as
\(Q\). This reference distribution \(Q\) summarises what good residuals
should follow given the predictors are known and fixed.

Probability distribution \(Q\) is actually a degenerate multivariate
distribution due to the fact that \(rank(\boldsymbol{R}) = n - 1 < n\).
This means the \(n\)th residual value will be known given the \(n - 1\)
residuals. To capture the exact characteristics of \(Q\), such as
moments, we can simulate a large numbers of \(\boldsymbol{\varepsilon}\)
and transform it to \(\boldsymbol{e}\). For simplicity, we replaced
\(\boldsymbol{R}\) with a diagonal matrix \(diag(\boldsymbol{R})\). The
resulting distribution for \(\boldsymbol{Q}\) is
\(N(\boldsymbol{0}_n, diag(\boldsymbol{R}\sigma^2))\).

If the model is misspecified, then the actual distribution of residuals
denoted as \(P\), will be different to \(Q\). For example, if the model
formula is correct, but the error \(\boldsymbol{\varepsilon}\) follows a
multivariate lognormal distribution, \(P\) will not be the same as \(Q\)
and we would expect a skewed empirical distribution for residuals. And
if there is an omitted variable problem, then the residual operator
obtained from the fitted model will not be the same as
\(\boldsymbol{R}\), and again, \(P\) will be different to \(Q\).

@@

Define a proper distance between distributions is usually easier than
define a proper distance between data plots. Given the actual residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance metric based on Kullback-Leibler divergence (ref here)
to quantify the difference between two distributions

\begin{align}
D &= log\left(1 + KL\right), \\
\label{eq:kl-1}
KL &= \int_{\mathbb{R}^{n}}log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{align}

\noindent where \(p(\boldsymbol{e})\) is the probability density
function for distribution \(P\), and \(q(\boldsymbol{e})\) is the
probability density function for distribution \(Q\).

This distance metric was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. For a linear regression model
that has omitted higher-order predictors \(\boldsymbol{Z}\) along with
coefficients \(\boldsymbol{\beta}_z\), and an error distribution
\(N(\boldsymbol{0}_n, \boldsymbol{V})\), \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}'))\).
Note that the variance-covariance matrix is replaced with its diagonal
analogy to ensure it is a full rank matrix.

@@

For non-normal error \(\boldsymbol{\varepsilon}\), the actual residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, equation \ref{eq:kl-2} will not be applicable to linear regression
model with non-normality issue.

To evaluate the Kullback-Leibler divergence of non-normal \(P\) from
\(Q\), we need to solve equation \ref{eq:kl-1}. However, since
\(\boldsymbol{e}\) is a linear transformation of non-normal random
variables, it is very common that the general form of \(P\) is unknown,
meaning that we can not easily compute \(p(\boldsymbol{e})\) using a
well-known probability density function.

Additionally, even if \(p(\boldsymbol{e})\) can be calculated for any
\(\boldsymbol{e} \in \mathbb{R}^n\), it will be very difficult to do
numerical integration over the \(n\) dimensional space, because \(n\)
could be potentially very large.

In order to evaluate equation \ref{eq:kl-1} in a practically computable
manner, the elements of \(\boldsymbol{e}\) are assumed to be independent
of each other. This assumption solves both of the issues mentioned
above. First, we no longer need to integrate over \(n\) random
variables. The result is now the sum of the Kullback-Leibler divergence
evaluated for each individual residual thanks for the independence
assumption. Second, it is not required to know the joint probability
density \(p(\boldsymbol{e})\) any more. Instead, the evaluation of
Kullback-Leibler divergence for a single residual relies on the
knowledge of \(p(e_i)\), where \(e_i\) is the \(i\)th residual for
\(i = 1, ..., n\). This is much easier to estimate through simulation.

The algorithm to compute equation \ref{eq:kl-1} starts from simulating
\(m\) sets of \(\boldsymbol{\varepsilon}\) according to the error
distribution. The simulated errors are stored in a matrix
\(\boldsymbol{A}\) with \(n\) rows and \(m\) columns. So each column of
\(\boldsymbol{A}\) is a set of realization values of
\(\boldsymbol{\varepsilon}\). Then, we can get \(m\) sets of
\(\boldsymbol{e}\) stored in the matrix \(\boldsymbol{B}\) by applying
the residual rotation operator
\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
density estimation (ref here) is applied on each row of \(B\) to
estimate \(p(e_i)\) for \(i = 1, ..., n\).

Since the Kullback-Leibler divergence can be viewed as the expectation
of the log-likelihood ratio between distribution \(P\) and distribution
\(Q\) evaluated on distribution \(P\), we can reuse the simulated
residuals in matrix \(\boldsymbol{B}\) to compute the expectation.
Therefore, the Kullback-Leibler divergence can be formulated as

\begin{align}
\label{eq:kl-3}
KL &= \sum_{i = 1}^{n} KL_i, \\
KL_i &= \frac{1}{m}\sum_{j = 1}^{m} log\frac{p(B_{ij})}{q(B_{ij})},
\end{align}

\noindent where \(KL_i\) is the Kullback-Leibler divergence for an
individual residual \(e_i\), \(B_{ij}\) is the \(i\)th row and \(j\)th
column entry of the matrix \(B\), \(q(.)\) is the normal density
function with mean zero and an assumed variance estimated as
\(\hat{\sigma^2} = var(flatten(B))\), and \(flatten(.)\) unfolds a
matrix into a vector.

\hypertarget{approximation-of-the-distance-metric-with-a-residual-plot}{%
\subsubsection{Approximation of the distance metric with a residual
plot}\label{approximation-of-the-distance-metric-with-a-residual-plot}}

In the previous sections, we have defined a distance metric for
quantifying the difference between the actual residual distribution and
an ideal reference distribution. This distance metric can only be
computed when the data generating process is known. In reality, we often
have no knowledge about the data generating process, otherwise we do not
need to fit a linear regression model in the first place.

What we proposed in this paper is a method to approximate this distance
with a residual plot. This is done by training computer vision models
with a large number of pairs of residual plots and distance. With the
approximated distance, we will be able to know how different the
underlying distribution of the residuals is from a good residual
distribution. This is meaningful way to check if a residual plot is a
good residual plot, and to know the strength of the visual signal
indicating model violations.

The approximated distance is not expected to be the same as the original
distance. This is largely because information embedded in a single
residual plot is limited and it may not be able to summarise the
characteristics of the residual distribution. For a given residual
distribution \(P\), we can generate many different residual plots. Some
of them share similar visual patterns, but some of them could be
visually very different from the rest, especially for models with small
\(n\). This suggests the approximated distance will vary depends on
whether the observed residual plot is representative or not.

\hypertarget{generation-of-simulated-training-data}{%
\subsection{Generation of simulated training
data}\label{generation-of-simulated-training-data}}

The computer vision models we developed in this study were trained with
synthetic data. This means the training data and the test data were
simulated from a known data generating process under different parameter
settings.

We have incorporated three types of residual departures of linear
regression model in this study, including non-linearity,
heteroskedasticity and non-normality.

\begin{align*}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= He_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= He_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \sqrt{\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^2}, \\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}_n, \boldsymbol{I}_n),
\end{align*}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(He_j(.)\) is the \(j\)th-order probabilist's Hermite polynomials
(ref), the \(\sqrt{(.)}\) and \((.)^2\) operator are element-wise
operators, \(\odot\) is the Hadamard product, and \(g(., k)\) is a
scaling function to enforce the support of the random vector to be
\([-k, k]^n\) defined as

@@

\hypertarget{different-configurations-of-the-model-formula}{%
\subsection{Different configurations of the model
formula}\label{different-configurations-of-the-model-formula}}

discuss different types of inputs and output choices, and why we choose
the current design

@@

@@

@@

>>>>

\newpage
\tableofcontents
\newpage

@@

There is a substantial need to develop an approach that alleviates
people's workload by automating repetitive tasks and providing
standardized results in a controlled environment. The large-scale
evaluation of lineups ihs impractical without the use of technology and
machines.

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatterplot diagnostics. \citet{wilkinson2005graph} expanded
on this work, introducing graph theoretic scagnostics, which encompassed
nine computable measures applied to planar proximity graphs. These
measures, including, but not limited to, ``Outlying'', ``Skinny'',
``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'', and
``Striated'' aimed to characterize outliers, shape, density, trend,
coherence and other characteristics of the data. While this approach has
been inspiring, there is a recognition \citep{buja2009statistical} that
it may not capture all the necessary visual features that differentiate
actual residual plots from null plots. A more promising alternative
entails enabling computers to learn the function for extracting visual
features from residual plots. Essentially, this means empowering
computers to discern the crucial visual features for residual
diagnostics and determining the method to extract them. Modern computer
vision models are well-suited for addressing this challenge.

@@

\hypertarget{different-possible-configurations-of-the-model-formula}{%
\subsection{Different possible configurations of the model
formula}\label{different-possible-configurations-of-the-model-formula}}

In addressing a problem, there exist multiple avenues for solutions.
Various configurations of the computer vision model formula can
effectively read residual plots, and these configurations can be
categorized based on two key components of the model formula: the input
and the output format.

\hypertarget{different-input-formats}{%
\subsubsection{Different input formats}\label{different-input-formats}}

In computer vision models, the input plays a role similar to predictors
in a linear regression model. The quality and relevance of the input
data greatly influence the model's capacity to generate insightful and
meaningful results. A straightforward approach involves feeding the
model a vector of residuals along with a vector of fitted values,
essentially providing all the necessary details for creating a residuals
vs fitted values plot. However, a drawback of this method is the dynamic
input size, as different fitted regression models may have varying
numbers of observations. In modern computer vision models created with
mainstream software like TensorFlow (ref here), the input is typically a
fixed-shape tensor. One solution is to pad the input vector with leading
or trailing zeros if the input tensor expects a longer vector. However,
this approach may fail if the input vector surpasses the designed
length. Another strategy is to summarize the residuals and fitted values
separately using histograms and utilize the counts as the input. By
controlling the number of bins in the histograms, it becomes possible to
provide a fixed-length input vector.

The primary advantage of using an image as input, as opposed to a vector
input format, relates to the development of sophisticated image
processing architectures over the years (see ref here). These
architectures effectively capture and summarize spatial information from
nearby pixels, which is less straightforward with vector input. The only
considerations are the image resolution and the aesthetics of the
residual plot. In general, higher resolution provides more information
to the model but comes with the trade-off of increased complexity and
greater difficulty in training. As for the aesthetics of the residual
plot, a practical solution is to consistently present residual plots in
the same style to the model. This implies that the model can not accept
arbitrary images as input but requires the use of the same preprocessing
software to convert provided residuals and fitted values into a
standardized style for the residual plot.

Other possible input formats include, but are not limited to, a pair of
residual plots, a triplet, and a lineup. Existing computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar (as seen in LeCun's 2005 paper). Applied to our
specific problem, we can define null plots of a fitted model to be
similar to each other, while considering actual residual plots to be
distinct from null plots of any fitted model. A triplet constitutes a
set of three images, denoted as \(image_1\), \(image_2\) and
\(image_3\). In computer vision models, triplets are often used to
predict whether \(image_2\) or \(image_3\) is more similar to
\(image_1\), proving particularly useful for establishing rankings
between samples. In the context of our problem, we can apply the same
criteria to define similarity between images. However, it's important to
note that these two approaches usually require additional considerations
regarding the loss function and, at times, non-standard training
processes due to shared weights between different convolutional blocks.

Presenting a lineup to a model and asking it to predict which residual
plot is the most different one aligns closely with the lineup protocol.
However, if the lineup consists of around 20 plots, similar to previous
human subject experiments (reference here), the resolution of the input
image could become quite large, posing challenges in training the model.
In fact, we experimented with this approach in a pilot study, and the
model's performance was suboptimal.

We did not explore all the mentioned input formats due to the
considerable costs associated with data preparation and model training.
Considering the implementation cost and the interpretability of the
model, we settled on the single residual plot input format.

\hypertarget{different-output-formats}{%
\subsubsection{Different output
formats}\label{different-output-formats}}

Given that the input is a single residual plot represented as a
fixed-resolution image, the computer vision model's output can take one
of two formats: binary or numeric. This choice determines whether the
model is a classification model or a regression model. The binary
outcome might indicates whether the input image is a null plot or not,
or whether the input image would be rejected in a visual test conducted
by humans. The latter option requires data from prior human subject
experiments, posing challenges in controlling the quality and quantity
of data due to variations in experimental settings across different
studies. Additionally, some visual inference experiments are unrelated
to linear regression models and residual plots, resulting in a limited
amount of available training data.

Alternatively, the output could be a meaningful numeric measure, such as
the average distance to good residual plots. This approach necessitates
defining a distance measure between residual plots, which may be a
non-trivial task. However, once a meaningful distance measure is
established, the computer vision model's prediction becomes an
interpretable value with tangible significance. Studies have
demonstrated that defining a proper distance between images can enhance
the matching accuracy in image search compared to a binary outcome model
(ref here).

The following section will propose the distance measure used for
training the computer vision model.

@@

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for evaluating residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image}. The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}. Furthermore, scagnostics mentioned in
Section \ref{introduction} could be used to construct distance metrics
for data plots comparison, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

@@

The distance metrics proposed in this paper takes into account the fact
that we try to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted model is
from a correctly specified model. For the classical normal linear
regression model, residuals are derived from the fitted values
\(\hat{\boldsymbol{y}}\) and observed values \(\boldsymbol{y}\). Suppose
the data generating process is known and the model is correctly
specified, by the Frisch-Waugh-Lowell theorem \citep{frisch1933partial},
residuals \(\boldsymbol{e}\) can also be written as a linear
transformation of the error \(\boldsymbol{\varepsilon}\) formulated as
\(\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator, \(\boldsymbol{X}\) is the design matrix,
\(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity matrix, and \(n\) is
the number of observations.

One of the assumptions of the classical normal linear regression model
is the error \(\boldsymbol{\varepsilon}\) follows a multivariate normal
distribution with zero mean and constant variance, i.e.,
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\).
It can be known that residuals \(\boldsymbol{e}\) also follow a certain
probability distribution transformed from the multivariate normal
distribution, which will be denoted as \(Q\). This reference
distribution \(Q\) summarises what good residuals should follow given
the predictors are known and fixed.

When fitting a linear regression model, the solver will force
\(\sum_{i=1}^{n} e_i = 0\), making any residual value to be a linear
combination of the remaining \(n - 1\) residuals. This effectively means
\(rank(\boldsymbol{R}) = n - 1 < n\) and \(Q\) becomes a degenerate
multivariate distribution. To capture the characteristics of \(Q\), such
as moments, we can simulate a large numbers of
\(\boldsymbol{\varepsilon}\) and transform it to \(\boldsymbol{e}\) to
get the empirical estimates. For simplicity, we replaced
\(\boldsymbol{R}\) with a full-rank diagonal matrix
\(diag(\boldsymbol{R})\), where \(diag(.)\) set the non-diagonal entries
of a matrix to zeros. The resulting distribution for \(\boldsymbol{Q}\)
is \(N(\boldsymbol{0}_n, diag(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different to \(Q\). For example, if
the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different to \(Q\)
because the residual operator obtained from the fitted model will not be
the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, the empirical residual distribution
will usually be skewed and has a long tail.

@@

Define a proper distance between distributions is usually easier than
define a proper distance between data plots. Given the actual residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance metric based on Kullback-Leibler divergence
\citep{kullback1951information} to quantify the difference between two
distributions

\begin{align}
\label{eq:kl-0}
D &= log\left(1 + KL\right), \\
\label{eq:kl-1}
KL &= \int_{\mathbb{R}^{n}}log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{align}

\noindent where \(p(.)\) is the probability density function for
distribution \(P\), and \(q(.)\) is the probability density function for
distribution \(Q\).

This distance metric was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2023plot} showed that,
for a classical normal linear regression model that omits a necessary
higher-order predictors \(\boldsymbol{Z}\), and incorrectly assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\),
\(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}'))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

@@

For non-normal error \(\boldsymbol{\varepsilon}\), the actual residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, equation \ref{eq:kl-2} given in \citet{li2023plot} will not be
applicable to models with non-normality violations.

To evaluate the Kullback-Leibler divergence of non-normal \(P\) from
\(Q\), the fallback is to solve equation \ref{eq:kl-1} numerically.
However, since \(\boldsymbol{e}\) is a linear transformation of
non-normal random variables, it is very common that the general form of
\(P\) is unknown, meaning that we can not easily compute
\(p(\boldsymbol{e})\) using a well-known probability density function.
Additionally, even if \(p(\boldsymbol{e})\) can be calculated for any
\(\boldsymbol{e} \in \mathbb{R}^n\), it will be very difficult to do
numerical integration over the \(n\) dimensional space, because \(n\)
could be potentially very large.

In order to evaluate equation \ref{eq:kl-1} in a practically computable
manner, the elements of \(\boldsymbol{e}\) are assumed to be independent
of each other. This assumption solves both of the issues mentioned
above. First, we no longer need to integrate over \(n\) random
variables. The result of equation \ref{eq:kl-1} is now the sum of the
Kullback-Leibler divergence evaluated for each individual residual
thanks for the independence assumption. Second, it is not required to
know the joint probability density \(p(\boldsymbol{e})\) any more.
Instead, the evaluation of Kullback-Leibler divergence for an individual
residual relies on the knowledge of the marginal density \(p_i(e_i)\),
where \(e_i\) is the \(i\)-th residual for \(i = 1, ..., n\). This is
much easier to estimate through simulation.

The algorithm for computing equation \ref{eq:kl-1} starts from
simulating \(m\) sets of \(\boldsymbol{\varepsilon}\) according to the
error distribution. The simulated errors are stored in a matrix
\(\boldsymbol{A}\) with \(n\) rows and \(m\) columns. So each column of
\(\boldsymbol{A}\) is a set of realization values of
\(\boldsymbol{\varepsilon}\). Then, we can get \(m\) sets of
\(\boldsymbol{e}\) stored in the matrix \(\boldsymbol{B}\) by applying
the residual rotation operator
\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
density estimation (KDE) with Gaussian kernel and optimal bandwidth
selected by the Sliverman's rule of thumb \citep{silverman2018density}
is applied on each row of \(B\) to estimate \(p_i(e_i)\) for
\(i = 1, ..., n\). The KDE computation is done by the \texttt{density}
function in R.

Since the Kullback-Leibler divergence can be viewed as the expectation
of the log-likelihood ratio between distribution \(P\) and distribution
\(Q\) evaluated on distribution \(P\), we can reuse the simulated
residuals in matrix \(\boldsymbol{B}\) to estimate the expectation by
the sample mean. With the independence assumption, for non-normal \(P\),
\(KL\) can be estimated by

\begin{align}
\label{eq:kl-3}
\hat{KL} &= \sum_{i = 1}^{n} \hat{KL_i}, \\
\hat{KL_i} &= \frac{1}{m}\sum_{j = 1}^{m} log\frac{\hat{p_i}(B_{ij})}{q(B_{ij})},
\end{align}

\noindent where \(\hat{KL_i}\) is the estimator of the Kullback-Leibler
divergence for an individual residual \(e_i\), \(B_{ij}\) is the
\(i\)-th row and \(j\)-th column entry of the matrix \(B\),
\(\hat{p_i}(.)\) is the kernel density estimator of \(p_i(.)\), \(q(.)\)
is the normal density function with mean zero and an assumed variance
estimated as
\(\hat{\sigma^2} = \sum_{b \in vec(B)}(b - \sum_{b \in vec(B)} b/nm)^2/(nm - 1)\),
and \(vec(.)\) is the vectorization operator which turns a
\(n \times m\) matrix into a \(nm \times 1\) column vector by stacking
the columns of the matrix on top of each other.

\hypertarget{approximation-of-the-distance-metric}{%
\subsubsection{Approximation of the distance
metric}\label{approximation-of-the-distance-metric}}

In the previous sections, we have defined a distance metric given in
equation \ref{eq:kl-0} for quantifying the difference between the actual
residual distribution \(P\) and an ideal reference distribution \(Q\).
You may have noticed that this distance metric can only be computed when
the data generating process is known. In reality, we often have no
knowledge about the data generating process, otherwise we do not need to
fit a linear regression model in the first place.

What we proposed in this paper is a method to approximate this distance
with a residual plot. Let \(D\) be the result of equation \ref{eq:kl-0}
indicating the extent of the model violations, and our estimator
\(\hat{D}\) is formulated as

\begin{equation}
\label{eq:d-approx}
\hat{D} = cv(plot_{h\times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
\end{equation}

\noindent where \(\boldsymbol{e}\) is a vector of residuals obtained
from the fitted model rather than a vector of random variables,
\(\hat{\boldsymbol{y}}\) is the fitted values also obtained from the
fitted model, \(plot(.)_{h \times w}\) is a plotting function that
generates a residuals vs fitted values plot with fixed aesthetic and
then saves it as an image with \(h \times w\) pixels and three colour
channels, and \(cv(.)\) is a computer vision model which takes an
\(h \times w\) image as input and predicts the distance.

With the approximated distance \(\hat{D}\), we will be able to know how
different the underlying distribution of the residuals is from a good
residual distribution. This is meaningful way to check if a residual
plot is a good residual plot, and to know the strength of the visual
signal embedded in the residual plot.

The approximated distance \(\hat{D}\) is not expected to be the same as
the original distance \(D\). This is largely because information
contained in a single residual plot is limited and it may not be able to
summarise all the characteristics of the residual distribution. For a
given residual distribution \(P\), we can generate many different
residual plots. Some of them share similar visual patterns, but some of
them could be visually very different from the rest, especially for
models with small \(n\). This suggests the error of the approximation
will vary depends on whether the observed residual plot is
representative or not.

\hypertarget{statistical-testing-based-on-the-approximated-distance}{%
\subsection{Statistical testing based on the approximated
distance}\label{statistical-testing-based-on-the-approximated-distance}}

\hypertarget{null-distribution-of-the-approximated-distance}{%
\subsubsection{Null distribution of the approximated
distance}\label{null-distribution-of-the-approximated-distance}}

\hypertarget{estimation-of-quantiles-of-the-null-distribution}{%
\subsubsection{Estimation of quantiles of the null
distribution}\label{estimation-of-quantiles-of-the-null-distribution}}

\hypertarget{bootstrapping-the-approximated-distance}{%
\subsubsection{Bootstrapping the approximated
distance}\label{bootstrapping-the-approximated-distance}}

\hypertarget{generation-of-training-data}{%
\subsection{Generation of training
data}\label{generation-of-training-data}}

\hypertarget{data-generating-process}{%
\subsubsection{Data generating process}\label{data-generating-process}}

While observational data is frequently employed in training models for
real-world applications, the data generating process of observational
data often remains unknown, making computation for our target variable
\(D\) unattainable. Consequently, the computer vision models developed
in this study were trained using synthetic data. This approach provides
us with precise label annotations. Additionally, it ensures a large and
diverse training dataset, as we have control over the data generating
process, and the simulation of the training data is relatively
cost-effective.

We have incorporated three types of residual departures of linear
regression model in the training data, including non-linearity,
heteroskedasticity and non-normality. All three departures can be
summarised by the data generating process formulated as

\begin{align}
\label{eq:data-sim}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= He_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= He_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \sqrt{\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^2},
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
predictors, \(He_j(.)\) is the \(j\)th-order probabilist's Hermite
polynomials \citep{hermite1864nouveau}, the \(\sqrt{(.)}\) and \((.)^2\)
operators are element-wise operators, \(\odot\) is the Hadamard product,
and \(g(., k)\) is a scaling function to enforce the support of the
random vector to be \([-k, k]^n\) defined as

@@

The residuals and fitted values of the fitted model is obtained by
regressing \(\boldsymbol{y}\) on \(\boldsymbol{x}_1\). This data
generating process is adopted from \citet{li2023plot} where it was used
to simulate residual plots with non-linearity and heteroskedasticity
visual patterns for human subject experiments.

In equation \ref{eq:data-sim}, predictor \(\boldsymbol{x}_1\) and
\(\boldsymbol{x}_2\) are independent of each other, so excluding
\(\boldsymbol{x}_2\) from the regression formula will not cause any
problem. However, \(\boldsymbol{z}\) and \(\boldsymbol{w}\) are
higher-order terms of \(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\). If
\(\beta_2 \neq 0\), the regression model will suffer from non-linearity
issues. Parameter \(j\) is a shape parameter controlling the number of
tuning points of the non-linearity pattern. Generally, greater values of
\(j\) will result in more tuning points.

Additionally, Scaling factor \(\boldsymbol{k}\) directly affects the
error distribution and it is correlated with \(\boldsymbol{x}_1\) and
\(\boldsymbol{x}_2\). If \(b \neq 0\) and
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\),
the constant variance assumption will be violated. Parameter \(a\) is a
shape parameter controlling the location of the smallest variance in a
residual plot.

Non-normality violations are introduced by defining a non-normal
distribution for \(\boldsymbol{\varepsilon}\). - distribution of x1 and
x2 - summary table of all the factor values - example plots for the
violations

\hypertarget{computation-of-scagnostics}{%
\subsubsection{Computation of
scagnostics}\label{computation-of-scagnostics}}

We have mentioned in Section \ref{introduction} that scagnostics
summarise some useful characteristics of the scatter plot. Although the
computer vision model will learn the feature extraction function during
the training process, it will be beneficial to utilize scagnostics to
help computer vision model making more accurate predictions as these
measure could provide additional useful information.

For each generated residual plot, four scagnostics, namely,
``Monotonic'', ``Sparse'', ``Splines'' and ``Striped'' are computed
using the \texttt{cassowaryr} R package (ref here). The computed
measures along with the number of observations of the fitted model are
provided as the second input of the computer vision model.

Other scagnostics are also informative, but unfortunately, they are
unavailable due to a fatal bug in the compiled C program of the
\texttt{interp} R package (ref here) that will crash the process in a
unpredictable manner. For reproducibility, we do not include these
scagnostics in the training data.

\hypertarget{crafting-a-balanced-training-set}{%
\subsubsection{Crafting a balanced training
set}\label{crafting-a-balanced-training-set}}

@@

\begin{itemize}
\tightlist
\item
  optimization (MSE loss function, with equation)
\item
  a diagram
\item
  explain the existence of removable layers
\item
  explain three different resolutions
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  software environment (tf and keras)
\item
  tuner (keras\_tuner)
\item
  80\% validation data, 20\% training data
\item
  100 trials (Bayesian optimizer)
\item
  monitor RMSE
\item
  early stopping (50 epochs)
\item
  learning rate reduction (10 epochs, 0.5 factor)
\item
  max 2000 epochs (no one reach)
\item
  restore best epoch of the best trial
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  RMSE for the test set
\item
  R\^{}2
\item
  Mean bias deviation to understand the overall bias
\item
  quantile loss to understand how well the model captures the entire
  distribution of D
\item
  Percentage of predictions within a tolerance interval (like 0.1)
\end{itemize}
